Gaussian variance estimation in simulated data sets

### 2018-11-09 analysis/gaussvarest.Rmd
wflow_publish("gaussvarest.Rmd")

This script implements the "Gaussian variance estimation" simulation
experiments in the paper. In particular, we compare the Mean Field
Variational Bayes method (MFVB) against SMASH in two scenarios.

@@

First simulation scenario
-------------------------

```{r simulation-1}
mse.mu.uneven.mfvb <- 0
mse.mu.uneven.haar <- 0
mse.mu.uneven.s8   <- 0

mse.sd.uneven.mfvb <- 0
mse.sd.uneven.haar <- 0
mse.sd.uneven.s8   <- 0
mse.sd.uneven.j    <- 0
mse.sd.uneven.s8.j <- 0

lmse.sd.uneven.mfvb <- 0
lmse.sd.uneven.haar <- 0
lmse.sd.uneven.s8   <- 0
lmse.sd.uneven.j    <- 0
lmse.sd.uneven.s8.j <- 0

# Repeat for each data set simulated in the first simulation scenario.
cat(sprintf("Running %d simulations for Simulation Scenario 1.\n",nsim1))
for (j in 1:nsim1) {
  cat(sprintf("%d ",j))

  # Create the simulated data set
  # -----------------------------
  set.seed(3*j)
  n     <- 500
  xOrig <- runif(n)
  set.seed(3*j)
  yOrig <- fTrue(xOrig) + sqrt(exp(loggTrue(xOrig)))*rnorm(n)

@@

  # Run SMASH
  # ---------
  x.mod <- unique(sort(xOrig))
  y.mod <- 0
  for(i in 1:length(x.mod))
    y.mod[i] <- median(yOrig[xOrig == x.mod[i]])

@@

  mse.mu.uneven.mfvb[j] <- mean((fhatMFVBgOrig - fTrue(xgOrig))^2)
  mse.mu.uneven.haar[j] <- mean((mu.est.inter - fTrue(xgOrig))^2)

  mse.sd.uneven.mfvb[j]=mean((sqrtghatMFVBgOrig-exp((loggTrue(xgOrig))/2))^2)
  mse.sd.uneven.haar[j]=mean((sqrt(var.est.inter)-exp((loggTrue(xgOrig))/2))^2)
  lmse.sd.uneven.mfvb[j]=mean((log(sqrtghatMFVBgOrig)-(loggTrue(xgOrig))/2)^2)
  lmse.sd.uneven.haar[j]=
    mean((log(sqrt(var.est.inter))-(loggTrue(xgOrig))/2)^2)

  mu.est       <- smash.gaus(y.final,filter.number=8,family="DaubLeAsymm")
  var.est      <- smash.gaus(y.final,v.est=TRUE,v.basis=TRUE,filter.number=8,
                             family="DaubLeAsymm")
  var.est.s8.j <- smash.gaus(y.final,v.est=TRUE,v.basis=TRUE,jash=TRUE,
                             filter.number=8,family="DaubLeAsymm")
  var.est.j    <- smash.gaus(y.final,v.est=TRUE,jash=TRUE)

  mu.est       <- mu.est[1:500]
  var.est      <- var.est[1:500]
  var.est.j    <- var.est.j[1:500]
  var.est.s8.j <- var.est.s8.j[1:500]

  mu.est.inter       <- approx(x.mod,mu.est,xgOrig,'linear')$y
  var.est.inter      <- approx(x.mod,var.est,xgOrig,'linear')$y
  var.est.inter.j    <- approx(x.mod,var.est.j,xgOrig,'linear')$y
  var.est.inter.s8.j <- approx(x.mod,var.est.s8.j,xgOrig,'linear')$y

  mse.mu.uneven.s8[j]=mean((mu.est.inter-fTrue(xgOrig))^2)
  mse.sd.uneven.s8[j]=mean((sqrt(var.est.inter)-exp((loggTrue(xgOrig))/2))^2)
  mse.sd.uneven.j[j] =mean((sqrt(var.est.inter.j)-exp((loggTrue(xgOrig))/2))^2)
  mse.sd.uneven.s8.j[j] =
    mean((sqrt(var.est.inter.s8.j) - exp((loggTrue(xgOrig))/2))^2)
  lmse.sd.uneven.s8[j]=mean((log(sqrt(var.est.inter))-(loggTrue(xgOrig))/2)^2)
  lmse.sd.uneven.j[j]=mean((log(sqrt(var.est.inter.j))-(loggTrue(xgOrig))/2)^2)
  lmse.sd.uneven.s8.j[j] =
    mean((log(sqrt(var.est.inter.s8.j))-(loggTrue(xgOrig))/2)^2)
}
cat("\n")
```

```
# SECOND SIMULATION SCENARIO
# --------------------------
mse.mu.even.mfvb <- 0
mse.mu.even.haar <- 0
mse.mu.even.s8   <- 0

mse.sd.even.mfvb <- 0
mse.sd.even.haar <- 0
mse.sd.even.s8   <- 0
mse.sd.even.j    <- 0
mse.sd.even.s8.j <- 0

lmse.sd.even.mfvb <- 0
lmse.sd.even.haar <- 0
lmse.sd.even.s8   <- 0
lmse.sd.even.j    <- 0
lmse.sd.even.s8.j <- 0

# Repeat for each data set simulated in the second simulation scenario.
cat(sprintf("Running %d simulations for Simulation Scenario 2.\n",nsim2))
for (j in 1:nsim2) {
  cat(sprintf("%d ",j))

@@

# PLOT MEAN & VARIANCE FUNCTIONS USED TO SIMULATE DATA
# ----------------------------------------------------
# Compare this plot against the one shown in Fig. 4 of the paper.
cat("Plotting mean and variance functions used to simulate data.\n")
xgrid <- (0:10000)/10000
plot(xgrid,fTrue(xgrid),type = "l",ylim = c(-5,5),ylab = "y",xlab = "X",
     lwd = 2)
lines(xgrid,fTrue(xgrid) + 2*sqrt(gTrue(xgrid)),col = "darkorange",lwd = 2)
lines(xgrid,fTrue(xgrid) - 2*sqrt(gTrue(xgrid)),col = "darkorange",lwd = 2)

>>>>

This script implements the "Gaussian variance estimation" simulation
experiments in the paper. In particular, we compare the Mean Field
Variational Bayes (MFVB) method against SMASH in two scenarios.

@@

Plot mean and variance functions used to simulate data
------------------------------------------------------

Compare this plot against the one shown in Fig. 4 of the paper.

```{r plot-mean-and-variance, fig.height=4, fig.width=5}
xgrid <- (0:10000)/10000
plot(xgrid,fTrue(xgrid),type = "l",ylim = c(-5,5),ylab = "y",xlab = "X",
     lwd = 2)
lines(xgrid,fTrue(xgrid) + 2*sqrt(gTrue(xgrid)),col = "darkorange",lwd = 2)
lines(xgrid,fTrue(xgrid) - 2*sqrt(gTrue(xgrid)),col = "darkorange",lwd = 2)
```

First simulation scenario: unevenly spaced data
-----------------------------------------------

In the first scenario, we simulated unevenly spaced data points, and
assessed accuracy by computing the mean of the squared errors (MSE)
evaluated at 201 equally spaced points.

```{r simulation-1-initialize, eval=FALSE}
mse.mu.uneven.mfvb  <- 0
mse.mu.uneven.smash <- 0
mse.sd.uneven.mfvb  <- 0
mse.sd.uneven.smash <- 0
```

Run the SMASH and MFVB methods for each simulated data set.

```{r simulation-1-main-loop, eval=FALSE} 
cat(sprintf("Running %d simulations: ",nsim1))
for (j in 1:nsim1) {
  cat(sprintf("%d ",j))

  # SIMULATE DATA
  set.seed(3*j)
  n     <- 500
  xOrig <- runif(n)
  set.seed(3*j)
  yOrig <- fTrue(xOrig) + sqrt(exp(loggTrue(xOrig)))*rnorm(n)

@@

  # RUN SMASH
  x.mod <- unique(sort(xOrig))
  y.mod <- 0
  for(i in 1:length(x.mod))
    y.mod[i] <- median(yOrig[xOrig == x.mod[i]])

@@

  mse.mu.uneven.mfvb[j] <- mean((fhatMFVBgOrig - fTrue(xgOrig))^2)
  mse.sd.uneven.mfvb[j] <- mean((sqrtghatMFVBgOrig-exp((loggTrue(xgOrig))/2))^2)

  mu.est  <- smash.gaus(y.final,filter.number=8,family="DaubLeAsymm")
  var.est <- smash.gaus(y.final,v.est=TRUE,v.basis=TRUE,filter.number=8,
                        family="DaubLeAsymm")

  mu.est  <- mu.est[1:500]
  var.est <- var.est[1:500]

  mse.mu.uneven.s8[j]=mean((mu.est.inter-fTrue(xgOrig))^2)
  mse.sd.uneven.s8[j]=mean((sqrt(var.est.inter)-exp((loggTrue(xgOrig))/2))^2)
}
cat("\n")
```

Second simulation scenario: evenly spaced points
------------------------------------------------

```{r, eval=FALSE}
mse.mu.even.mfvb <- 0
mse.mu.even.s8   <- 0

mse.sd.even.mfvb <- 0
mse.sd.even.s8   <- 0
lmse.sd.even.mfvb <- 0
lmse.sd.even.s8   <- 0

# Repeat for each data set simulated in the second simulation scenario.
cat(sprintf("Running %d simulations: ",nsim2))
for (j in 1:nsim2) {
  cat(sprintf("%d ",j))

@@
